---
title: "Final Project"
author: "Joe Opitz & Quinn Springer"
date: "2025-04-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
dfr <- read.csv("C:\\Users\\jopitz\\OneDrive - Montana Tech\\STAT 491 - Intro to Data Science\\Stock-Market\\S&P.csv", header = TRUE)
install.packages("dplyr");library(dplyr)
install.packages("ggplot2"); library(ggplot2)
library(rpart)
install.packages("rpart.plot");library(rpart.plot)
```

## We will start by preparing our S&P 500 data

```{r}
# View the first 5 rows of the data
head(dfr, 5)
```

We will mostly be using Adj.Close, or features created based off of Adj.Close.

```{r}
# Add index field
n <- dim(dfr)[1]
dfr$index <- c(1:n)

# Check for null values
sum(is.na(dfr))
```
No missing values.

### Outlier detection
```{r}
# Calculate percentage returns for consistency
dfr$Returns <- c(0,diff(log(dfr$Adj.Close)))

# Compute outlier threshold
mu.return <- mean(dfr$Returns)
sd.return <- sd(dfr$Returns)
outlier_threshold.pos <- mu.return + 3 *sd.return
outlier_threshold.neg <- mu.return - 3 *sd.return

# Detect outliers
SP_outliers <- dfr[which(dfr$Returns < outlier_threshold.neg |
                           dfr$Returns > outlier_threshold.pos),]

SP_outliers
```

#### We have a few outliers, so we will flag them for the models to see

```{r}
# Flag outliers
dfr$Outliers <- ifelse(dfr$Returns < outlier_threshold.neg |
                           dfr$Returns > outlier_threshold.pos, 1, 0)
```

```{r}
# Create lag-based log returns
for (k in 0:5) {
  dfr[[paste0("Lag_", k)]] <- dplyr::lag(dfr$Returns, k)
}
```


```{r}
head(dfr)
```

```{r}
# We will now have to remove Lag_0 as it is the same as Adj.Close
dfr$Lag_0 <- NULL
```

## We will now begin our EDA step

```{r}
# Change date to time-series data
dfr$Date <- as.Date(dfr$Date)

# Create plot
ggplot(dfr, aes(x = Date, y = Adj.Close)) +
  geom_line() +
  labs(title = "S&P 500 Over Time", x = "Date", y = "Adjusted Close") +
  theme_minimal()
```


```{r}
# Bin return values to visualize amount of increasing/decreasing days
dfr$Returns.binned <- cut(dfr$Returns, breaks = c(-1, 0, 1), labels = c("Decreasing", "Increasing"))

# Create plots to demonstrate amount of days increasing versus decreasing
ggplot(dfr, aes(Returns.binned)) + geom_bar()
```

We can see that there is about as many days where the price increases versus decreases, with slightly more increasing days

```{r}
# Create histogram of daily returns
ggplot(dfr, aes(Returns)) + geom_histogram(color = "black")
```

Seems to be fairly normally distributed with a couple of outliers, that we have already detected in a previous step 

## Create train/test split on data
```{r}
set.seed(406)

# Prepare for partition
n <- dim(dfr)[1]
train_ind <- runif(n) < 0.75

# Create train test split
df_train <- dfr[train_ind,]
df_test <- dfr[!train_ind,]
```

# Come back to and validate


## We will start off with a Multiple Linear Regression model
```{r}
# Create model
lm <- lm(Returns ~ Lag_1, data = df_train)

# View summary
summary(lm)
```

Based on these results, we are able to tell that the model is not doing very well. Which is too be expected for a project like this we will need to have much more powerful models than a simple linear regression model.

### Create a Multiple linear regression model
```{r}
# Create model
lm_multi <- lm(Returns ~ Lag_1 + Lag_2 + Lag_3 + Lag_4 + Lag_5, data = df_train)

# View model summary
summary(lm_multi)
```

As stated earlier, any sort of a linear regression model will be extremely difficult to predict the price of a stock. So, there is no surpise that the model is still not performing well. We will be begin to move towards some more advanced models.

## Try a cart model to predict whether or not the model will increase or decrease
```{r}
# Change categorical features
df_train$Returns.binned <- as.factor(df_train$Returns.binned)

CART <- rpart(Returns.binned ~ Lag_1 + Lag_2 + Lag_3  + Lag_4 + Volume, 
              data = df_train, method = "class")

ypreds <- predict(CART, newdata = df_test, type = "class")

t1 <- table(df_test$Returns.binned, ypreds)
t1 <- addmargins(t1)
t1
```

